Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-07 Dail_Mad_ext.tex
new draft intro

A simple state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a deterministic population growth model, $\mu(N_{t-1})$,
coupled with a stochastic model allowing
for random process variation. %, which we denote $\grave{o}_t$.
The deterministic model may be density-dependent, for example
logistic, or it might be density-independent, as in the case of
exponential growth, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$
is the intrinsic rate of increase. The full model can now be written
as:
%As an example, under
%density-independent exponential population growth, the
%model could be written as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
%    \mu_t &= N_{t-1}\exp(r) \quad \text{for} \; t=2,\hdots,T
    N_1 &= X_1 \label{eq:ss1a} \\
    N_t &= \mu(N_{t-1}) + \grave{o}_{t-1} \quad \text{for} \; t=2,\hdots,T \label{eq:ss1b} \\
    X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \; t=1,\hdots,T \label{eq:ss1c}
  \end{align}
\end{subequations}
%{\flushleft where} $\mu(N_{t-1})$ is the deterministic population
%growth model. %One might alternatively consider density-dependent models
%for $\mu(N_{t-1})$, and this flexibility is an appealing feature of
%state-space models.
In classical state-space models, the two sets of random effects %In most classical state-space models both sets of parameters
are assumed to be i.i.d Gaussian deviates:
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$.
Note that Eq~\ref{eq:ss1a} indicates that no process variation is
present in year 1. This assumption is often made
due to parameter identifiability problems, which we will describe
shortly. First, we turn our attention to the models for the random
effects.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
are the most widespread approach for modeling population dynamics
using time-series data, several problems are evident. Namely, (1)
time-series data and the underlying abundance parameters of interest
are typically integer-valued, raising concerns about the use of
Gaussian distribution for the random effects, (2) the model for
observatoin error has little biological basis, (3)
spatial variation in abundance is not allowed, and (4) some of the
paramters of the model are not estimable. We briefly discuss each of
these points before describing a general approach to resolving these
shortcomings.

These assumptions are motivated by
convenience rather than biology. Specifically, the Gaussian assumption
allows for parameter estimation using the Kalman filter
\citep{dennis_etal:2006}, which is much more computationally efficient
than estimation methods when random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}. The problem with this is that it
allows for negative values of the state-variable, which is
inconsistent with the observed data. The fact that both abundance ($N_t$) and the observed data ($X_t$) are
typically non-negative integers suggests that it is inappropriate to
model process variation and observation error using Gaussian
distributions. Nonetheless, efforts have been made to adhere to the
Gaussian assumptions by tranforming abundance to
density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
area surveyed. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with some small number, although the
effect of this is rarely discussed.

Another problem with standard state-space models is that the
observation error has no clear interpretation. The use of a mean zero
Gaussian distribution implies that the observed data $X_t$ will be
higher than $N_t$ as often as it is lower than $N_t$. It is hard
to identify a mechanism that would cause such errors. A more likely
form of observation error, and one that has been recognized for well
over a century, results from failing to detect individuals that are
present. Imperfect detection may be attributable to
characteristics of the species under study, such as its elusiveness,
or to the failings of the ecologist collecting the data in the field.
Although a vast number of methods have been devised for accounting for
this form of observation error, rarely have these methods been
integrated into state-space models \citep[but
see][]{buckland_etal:2004}.

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$ or
$X_1$; however, such distributions are needed to fully specify the
model. Indeed, a fully-specified state-space model requires the
specification of three probability distributions, which we denote
using bracket notation:
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \\ \label{eq:ss2b}
    [N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\ \label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters governing
initial abundance, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. In our first example, Eq.~\ref{eq:ss1},
But what should these distributions be, and how could
parameters of the distributions be
estimated since there is only a single observation available? In fact,
they cannotThe reason they were omitted is that it is not possible to
estimate the parameters of the i
However, a completSpecifically, the Markovian
nature of the model implies the following likelihood
The first term is the probability density for the first observation in
the time series, $X_1$, yet it is intuitively not possible to estimate the
parameter(s) of this distribution using only a single
observation.
%This problem has resulted in several papers with
%interesting names like ``multi-modal likelihoods and ridges etc...''
Two options have been proposed handling the identifiability
problem. First, one can assume that the first observation in the time
series is made without error. Second, one can assume that the population is at equilibrium
so that that the x can be replaced with the equilibrium
distribution. However, assuming equilibrium defeats the objective of
many studies of population dynamics, namely determining if and why a
population is at equilibrium.

@@

Several extensions of state-space models have been proposed to
overcome the limitations described
above. \citet{devalpine_hastings:2002} and \citet{kery_etal:2009}
described methods for fitting models with non-Gaussian distributions
for the process and observation errors. Observation models with more
intuitive interpretations, such as those that explicitly model
detection probability, have been proposed by
\citet{kery_etal:2009}. \citet{lele_etal:1998} and
\citep{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, the work by
\citet{kery_etal:2009} is unique in that it addressed multiple limitations
simultaneously. However, their models did not include any serial
depedence, which is a hallmark of population models. This limits the
utility of their model for making inferences about explicit population
processes. To do so, recent work has sought to combine multiple
sources of information, such as counts and mark-recapture data, to
allow for inferences about population processes such as mortality and
survival, rather than just population growth \citep{besbeas_etal:2002,
  buckland_etal:2004, schaub_etal:2007}. Although these integrated
population models may be viewed as the gold standard in state-space
modeling, ecologists are not always so fortunate to have direct
information about vital rates, especially when inferences are required
at large spatial scales. Rather, count data are much more common, and
are produced by many of the largest monitoring programs in the world,
such as the North American Breeding Bird Survey
\citep[BBS;][]{robbins_etal:1986}.

In this paper, we focus on the model of
\citet{dail_madsen:2011} %developed a model (henceforth the DM model)
that simultaneous resolves each of the abovementioned problems with
traditional state-space models, and is designed for simple count
data. The model (1) recognizes
the discrete nature of count data and the underlying true abundance
state, (2) includes a realistic observation model with explicit
detection probability parameters (3) has estimable parameters,
including those describing the initial state at time $t=1$, and (4)
allows for inference about both temporal and spatial variation in
abundance.

\begin{comment}
  In some respects, the DM model can be viewed as a specific
  manifestation of the general state-space model formulated by
  \citep{buckland_etal:2004}. However, their model was designed to
  allow for inference about demographic parameters, which will
  typically require mark-recapture data. Although this may be viewed
  as the gold standard in many studies of wildlife populations, in
  this paper, we focus on the analysis of count data, which is the
  typical source of information used in state-state modeling
  applications, and one of the most widely available forms of
  ecological data---it is also the data required by the DM model.
\end{comment}

In the following section, we describe the DM model in its original
form, and we explain how it resolves each of the deficieinces with
standard state-space models. In Section~\ref{sec:ext}, we extend the
model to accomodate classical models of population growth and to
handle several features common to ecological
time-series. Specifically, we describe methods for accomodating excess
zeros and nuisance variables such as random observer effects. Both
frquentist and Bayesian methods of inference are discussed, and code
for fitting models is presented in the appendices. The Bayesian method
is attractive in that it can accomodate prior information about
detection parameters, which is available for many species, yet not
collected as part of many existing monitoring programs. In
Section~\ref{sec:app}, we evaluate the performance of the model
extensions using a simulation study and by analyzing data from the
North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}. The overarching aim
of  the paper is to provide ecologists with a flexible and accessible means of
adressing important questions related to the variaion of abundance in
space and time.
believe these extensions will increase the utility of the model for
application to existing datasets such the Breeding Bird Survey.

@@

The DM model is an extension of the $N$-mixture model
\citep{royle:2004biom}, which allows for inference about spatial
variation in abundance when individuals cannot be detected with
certainty. To estimate both abundance parameters and parameters of the
detection process, the model uses replicate observations at each site,
which are collected during sufficiently short time intervals such that
the population can safely be assumed to be closed with respect to
births, deaths, and movement. The DM model relaxes this closure
assumption and includes explicit parameters
describing population change over time.

>>>>

A simple state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a deterministic population growth model, $\mu(N_{t-1})$,
coupled with a stochastic model allowing
for random process variation. %, which we denote $\grave{o}_t$.
The deterministic model may be density-dependent, for example
logistic, or it might be density-independent, as in the case of
exponential growth, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$
is the intrinsic rate of increase. The full model can now be written
as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \label{eq:ss1a} \\
    N_t &= \mu(N_{t-1}) + \grave{o}_{t-1} \quad \text{for} \; t=2,\hdots,T \label{eq:ss1b} \\
    X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \; t=1,\hdots,T \label{eq:ss1c}
  \end{align}
\end{subequations}
where $\grave{o}_t$ is the random effect allowing for process
variation unaccounted for by the deterministic model. In classical
state-space models, the two sets of random effects
are assumed to be i.i.d Gaussian deviates:
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. It is also standard practice to
act as though there is no process variation associated with $N_1$, as
indicated by Eq~\ref{eq:ss1a}.
% indicates that no process variation is
%present in year 1. This assumption is often made
%due to parameter identifiability problems, which we will describe
%shortly. First, we turn our attention to the models for the random
%effects.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
are the most widespread approach for modeling population dynamics
using time-series data, several problems are evident. Namely, (1)
time-series data and the underlying abundance parameters of interest
are typically integer-valued, as in the case of count data, raising
concerns about the use of Gaussian distribution for the random
effects, (2) the model for observation error has little biological basis, (3)
spatial variation in abundance is not allowed, and (4) some of the
paramters of the model are not estimable. We briefly discuss each of
these points before describing a general approach to resolving these
shortcomings.

Use of the Gaussian distribution for modeling random process variation
and observation error is motivated by convenience rather than
biology. Specifically, the Gaussian assumptions
allow for parameter estimation using the Kalman filter
\citep{dennis_etal:2006}, which is much more computationally efficient
than estimation methods when random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}. The problem with this is that it
allows for negative values of the state-variable, which is
inconsistent with the observed data.
%The fact that both abundance ($N_t$) and the observed data ($X_t$) are
%typically non-negative integers suggests that it is inappropriate to
%model process variation and observation error using Gaussian
%distributions.
Nonetheless, efforts have been made to adhere to the
Gaussian assumptions by tranforming abundance to
density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
area surveyed. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with some small number, although the
effect of this is rarely discussed.

Another problem with standard state-space models is that the
observation error has no clear interpretation. The use of a mean zero
Gaussian distribution implies that $X_t$ will be
higher than $N_t$ as often as it is lower than $N_t$. It is hard
to identify a mechanism that would cause such symmetric errors. A more likely
form of observation error, and one that has been recognized for well
over a century, results from failing to detect individuals that are
present. Imperfect detection may be attributable to
characteristics of the species under study, such as its elusiveness,
or to the failings of the ecologist collecting the data in the field.
Although a vast number of methods have been devised for accounting for
this form of observation error, rarely have these methods been
integrated into state-space models \citep[but
see][]{buckland_etal:2004}.

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
however, $N_1$ is a random variable and hence there is uncertainty
that should be accounted for.
%are needed to fully specify the model.
To be more specific, a fully-specified state-space model requires at least
three probability distributions, which we represent
using bracket notation:
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \\ \label{eq:ss2b}
    [N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\ \label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters for the
initial state, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. The three lines of general model
described by Eq.~\ref{eq:ss2} correspond to the three equations shown
for the specific example in Eq~\ref{eq:ss1}.  However, in this case,
we allow $N_1$ to be a random variable. But what should its
distribution, $[N_1]$ be? And how could $\bm{\theta}$ be
estimated since there is only a single observation available?
One approach to estimating the $\bm{\theta}$ is to assume that the
population is in equlibrium, i.e. that the expected value of $N_t$ is
constant through time. Althought this makes the parameters
identifiable, assuming equlibrium defeats the objective of many
studies of population dynamics, namely determining why a population
varies over time.

\begin{comment}
  However, a completSpecifically, the Markovian nature of the model
  implies the following likelihood The first term is the probability
  density for the first observation in the time series, $X_1$, yet it
  is intuitively not possible to estimate the parameter(s) of this
  distribution using only a single observation.
  % This problem has resulted in several papers with interesting names
  % like ``multi-modal likelihoods and ridges etc...''
  Two options have been proposed handling the identifiability
  problem. First, one can assume that the first observation in the
  time series is made without error. Second, one can assume that the
  population is at equilibrium so that that the x can be replaced with
  the equilibrium distribution. However, assuming equilibrium defeats
  the objective of many studies of population dynamics, namely
  determining if and why a population is at equilibrium.
\end{comment}

@@

Several extensions of state-space models have been proposed to
overcome the limitations described
above. \citet{devalpine_hastings:2002} and \citet{kery_etal:2009}
described methods for fitting models with non-Gaussian distributions
for the process and observation errors. Observation models with more
intuitive interpretations, such as those that explicitly model
detection probability, have been proposed by
\citet{kery_etal:2009}. \citet{lele_etal:1998} and
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, the work by
\citet{kery_etal:2009} is unique in that it addressed multiple limitations
simultaneously. However, their model did not include serial
depedence, which is a hallmark of population models. This limits the
utility of their model for making inferences about explicit population
processes. In contrast, recent work has sought to use state-space
models fot inference about population processes such as mortality and
recruitment. To do so, methods have been developed to combine multiple
sources of information, such as counts and mark-recapture data
\citep{besbeas_etal:2002, buckland_etal:2004,
  schaub_etal:2007}. Although these integrated
population models may be viewed as the gold standard in state-space
modeling, ecologists are not always so fortunate to have direct
information about vital rates, especially at large spatial scales.
Rather, count data are much more common, and
are produced by many of the largest monitoring programs in the world,
such as the North American Breeding Bird Survey
\citep[BBS;][]{robbins_etal:1986}.

In this paper, we focus on the model of
\citet{dail_madsen:2011} %developed a model (henceforth the DM model)
that simultaneous resolves each of the abovementioned problems with
traditional state-space models, and is designed for simple count
data.
\begin{comment}
  The model (1) recognizes
  the discrete nature of count data and the underlying true abundance
  state, (2) includes a realistic observation model with explicit
  detection probability parameters (3) has estimable parameters,
  including those describing the initial state at time $t=1$, and (4)
  allows for inference about both temporal and spatial variation in
  abundance.

  In some respects, the DM model can be viewed as a specific
  manifestation of the general state-space model formulated by
  \citep{buckland_etal:2004}. However, their model was designed to
  allow for inference about demographic parameters, which will
  typically require mark-recapture data. Although this may be viewed
  as the gold standard in many studies of wildlife populations, in
  this paper, we focus on the analysis of count data, which is the
  typical source of information used in state-state modeling
  applications, and one of the most widely available forms of
  ecological data---it is also the data required by the DM model.
\end{comment}
In the following section, we describe the DM model in its original
form and explain how it resolves each of the deficieinces with
standard state-space models. In Section~\ref{sec:ext}, we extend the
model to accomodate classical models of population growth and to
handle several features common to ecological
time-series. Specifically, we describe methods for accomodating excess
zeros and nuisance variables such as random observer effects. Both
frquentist and Bayesian methods of inference are discussed, and code
for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accomodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
In Section~\ref{sec:app}, we evaluate the performance of the model
extensions using a simulation study and by analyzing data from the
North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}. The overarching aim
of  the paper is to provide ecologists with flexible and accessible means of
adressing important questions related to the variaion of abundance in
space and time.
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.

@@

The DM model is an extension of the $N$-mixture model
\citep{royle:2004biom}, which allows for inference about spatial
variation in abundance when individuals cannot be detected with
certainty. To estimate both abundance parameters and parameters of the
detection process, the original $N$-mixture model uses replicate
observations at each site, which are collected during sufficiently
short time intervals such that the population can safely be assumed to
be closed with respect to births, deaths, and movement. The DM model
relaxes this closure assumption and includes explicit parameters
describing population change over time.
